---
title: "R Notebook"
output: html_notebook
---

```{r}

library(data.table)
library(dplyr)
library(readr)
library(tidyr)
# Load datasets 
outpatient <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/OutpatientVisit.csv")
patient <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/Patient.csv")
clinic <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/Clinic.csv")
staff <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/Staff.csv")
mortality <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/Mortality.csv")
icd_codes <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/ICDCodes.csv")
disease_map <- fread("/Users/swetha/Desktop/high performance computing/healthcare-data/DiseaseMap.csv")
```

```{r}
#1. Are men more likely to die than women in this group of patients? Assume people without a date of death in the mortality table are still alive.

# Merge patient data with mortality data
patient_mortality <- merge(patient, mortality, by = "PatientID", all.x = TRUE)

# Add a flag for death (1 = Deceased, 0 = Alive)
patient_mortality[, IsDeceased := !is.na(DateOfDeath)]

# Group by gender and calculate mortality rate
mortality_rate <- patient_mortality[, .(
  TotalPatients = .N,
  DeceasedPatients = sum(IsDeceased, na.rm = TRUE),
  MortalityRate = sum(IsDeceased, na.rm = TRUE) / .N * 100
), by = Gender]

print(mortality_rate)

```

# Yes, men are slightly more likely to die than women in this group of patients. The calculated mortality rate for men is 35.95%, while for women it is slightly lower at 35.11%. Although the difference is small, it indicates a marginally higher likelihood of death for male patients in this analysis. 

```{r}
#2. Are there patterns in the disease groups across gender. For every patient with at least one outpatient visit, identify if they have been diagnosed with any of the 22 conditions listed in the diseaseMap table at any time point. You will need to consider all three ICD columns in the outpatientVisit file (not just one). Create a table with the rate of disease for each condition for men, women, and all. It should look like this, where the XX% is the percent with the condition.

# Join outpatient visits with patient information to add gender
outpatient_gender <- outpatient %>%
  left_join(patient, by = "PatientID")

# Reshape outpatient visits to long format (one row per ICD code)
outpatient_diseases <- outpatient_gender %>%
  pivot_longer(cols = starts_with("ICD10"), names_to = "ICD_column", values_to = "ICD_code") %>%
  # Join ICD codes to get ICD description
  left_join(icd_codes, by = c("ICD_code" = "ICD10")) %>%
  # Join disease map to get the disease condition
  left_join(disease_map, by = c("ICD_code" = "ICD10"))

# Create a list of diseases to filter by
disease_list <- disease_map$Condition

# Filter data to only include records where the disease condition is in the disease list
outpatient_diseases_filtered <- outpatient_diseases %>%
  filter(Condition %in% disease_list) %>%
  distinct(PatientID, Condition, Gender)

# Ensure Gender column has no empty or missing values
outpatient_diseases_filtered <- outpatient_diseases_filtered %>%
  mutate(Gender = ifelse(is.na(Gender) | Gender == "", "Unknown", Gender))

# Group and calculate patient counts by Condition and Gender
disease_gender_rate <- outpatient_diseases_filtered %>%
  group_by(Condition, Gender) %>%
  summarise(
    Total_Patients = n_distinct(PatientID),
    .groups = "drop"
  ) %>%
  # Pivot the data to get separate columns for each gender and 'All'
  pivot_wider(
    names_from = Gender,
    values_from = Total_Patients,
    values_fill = list(Total_Patients = 0)
  ) %>%
  # Add a column for the total number of patients
  mutate(
    All = rowSums(select(., -Condition), na.rm = TRUE),
    Total_All = sum(All)
  ) %>%
  # Calculate the percentage for each condition for each gender and 'All'
  mutate(
    Men_Percent = ifelse(!is.na(male), (male / Total_All) * 100, 0),
    Women_Percent = ifelse(!is.na(female), (female / Total_All) * 100, 0),
    All_Percent = (All / Total_All) * 100
  )

# Final table with the disease rates for men, women, and all
final_disease_rate <- disease_gender_rate %>%
  select(Condition, Men_Percent, Women_Percent, All_Percent)

# View the final result
print(final_disease_rate)
```


```{r}
library(testthat)
test_that("summaries populated", expect_equal(nrow(final_disease_rate), 22))
```



#Analysis of the 22 conditions listed in the `diseaseMap` table reveals notable patterns across genders for patients with at least one outpatient visit. Hypertension was the most prevalent condition, affecting slightly more men (11.12%) than women (10.52%). Obesity and depression were more common among women, with obesity affecting 6.77% of women compared to 4.83% of men, and depression at 4.58% for women versus 2.91% for men. Alcohol and drug-related conditions were more prevalent in men, with alcohol at 2.77% for men and 2.86% for women, while drug-related conditions were at 1.33% for men and 1.50% for women. Diabetes, both with and without complications, showed higher prevalence among women, and conditions like stroke and myocardial infarction had slightly higher rates in men. These patterns emphasize the need for gender-sensitive healthcare interventions to address these disparities effectively.

```{r}
#Calculate the mortality rate for every year between 2005 and 2018. Is it generally increasing, or decreasing? Assume patients are only at risk of death as of their first visit (in the outpatient Visit file). Once they have died, they are no longer at risk in subsequent year.

# Identify the first outpatient visit for each patient
first_visit <- outpatient %>%
  group_by(PatientID) %>%
  summarise(FirstVisitDate = min(VisitDate), .groups = "drop")

# Merge with mortality data to get the date of death for each patient
data_mortality <- mortality %>%
  left_join(patient, by = "PatientID") %>%
  select(PatientID, DateOfDeath)

# Merge first visit and mortality data to determine at-risk period
data_patient <- first_visit %>%
  left_join(data_mortality, by = "PatientID") %>%
  mutate(
    DateOfDeath = as.Date(DateOfDeath),
    FirstVisitDate = as.Date(FirstVisitDate)
  )

# Calculate mortality rates for each year between 2005 and 2018
mortality_rate_year <- data_patient %>%
  filter(!is.na(FirstVisitDate)) %>%
# For each year of first visit, we need to calculate the mortality rate
  mutate(YearOfFirstVisit = year(FirstVisitDate)) %>%
  filter(YearOfFirstVisit >= 2005 & YearOfFirstVisit <= 2018) %>%
  group_by(YearOfFirstVisit) %>%
  summarise(
    TotalPatients = n_distinct(PatientID),
    TotalDeaths = sum(!is.na(DateOfDeath) & year(DateOfDeath) == YearOfFirstVisit),
    MortalityRate = (TotalDeaths / TotalPatients) * 100,
    .groups = "drop"
  )

# View the mortality rate by year
print(mortality_rate_year)

```


```{r}

get_mortality_rate <- function(year) {
  rate <- mortality_rate_year %>%
    filter(YearOfFirstVisit == year) %>%
    pull(MortalityRate)
  if (length(rate) == 0) return(NA)  
  return(rate)
}        

# Test mortality rates for all years between 2005 and 2018
test_that("Mortality rates are correct for each year", {
  for (year in 2005:2018) {
    calculated_rate <- get_mortality_rate(year)
    expected_rate <- mortality_rate_year %>%
      filter(YearOfFirstVisit == year) %>%
      pull(MortalityRate)
    expect_equal(calculated_rate, expected_rate)  
  }
})

```



#The mortality rate between 2005 and 2018 fluctuates rather than showing a clear increasing or decreasing trend. In 2005, the mortality rate was 3.96%, peaking at 5.77% in 2006 and fluctuating across the years, with notable rates such as 5.89% in 2016 and a drop to 3.12% in 2017. This variability suggests that no consistent pattern exists, and the changes may be influenced by factors like variations in patient demographics, healthcare access, or sample sizes across different years.

